Mathematical Language Processing \\ Project

### 2014-06-19 paper.tex
…

To build such a system, a labeled text corpus that annotates identifiers and
their definition is desirable. At the project start, such a corpus was not
available. 
We assume that the probability to find correct identifier definition tuples is
part-of-speech (POS) tagged based distance metrics.
Our motivation for our choice is that readers usually extract identifier 
definitions from context that is given by the surrounding text.
In combination with the Stratosphere system \todo{[HSC: NEED REFERENCE]}
that is capable of scalable data processing,  we rapidly generate a large
amount of definition relation candidates with only minimal implementation
overhead for the parallelization.

We chose the Wikipedia text corpus as a target because of two facts. First,
most articles make use of \texttt{<math/>} tags (\emph{texvc} as an input
language) for formulae.  The identification of these tags are easy to find, 
and from the MathML output, it is easy to extract the identifiers.
Second, the text is already annotated with mark-up.
Particularly, hyperlinks to other articles within Wikipedia are of
interest as they typically wrap around any number of words and indicate that
these in combination are relevant in the given context or (respectively) sentence.

@@

At first, we implemented a simple identifier definition extractor that is
based on a set of static patterns. As this is a fairly robust approach and easy
to implement, it serves as a good reference point in terms of
performance. The following patterns are being used to discover description
terms: \todo{[HSC: here you should refer to the table]}

\begin{table}
\vspace{-5pt}
	\begin{center}
		\begin{tabular}{| p{9.3cm} |}
			\hline
			Pattern \\
			\hline
			\texttt{\emph{<description>} \emph{<identifier>}} \\
			\texttt{\emph{<identifier>} is \emph{<description>}} \\
			\texttt{\emph{<identifier>} is the \emph{<description>}} \\
			\texttt{let \emph{<identifier>} be the \emph{<description>}} \\
			\texttt{\emph{<description>} is|are denoted by \emph{<identifier>}} \\
			\texttt{\emph{<identifier>} denotes */DT \emph{<description>}} \\
			\hline
		\end{tabular}
	\end{center}
\caption{Implemented static patterns}
\vspace{-5pt}
\end{table}

Note that in the last pattern we used a POS tag.  Determiners not only contain 
articles, but also quantifiers and distributives.
To increase readability, this list have been shortened to `*/DT'. 
\todo[inline]{[HSC: Explain */DT in detail.] describe the process: extraction of identifiers, 
how to identify the description? Looking for space or using POS tags?}

@@

Extracting relevant identifiers from the article rely on the assumption that
the author will use \texttt{<math/>} tags for all formulae. This said, a
formula that is written within the text cannot be recognized, and therefore
can never be extracted.

The fact that we can estimate all relevant identifiers for an article (see
Section \ref{ir}), combined with some common assumptions about the relation,
can be exploited to largely reduce the set of candidates that need to be ranked.
Please note that this reduction is essential for retrieving the correct
relations. Otherwise almost any word would be ranked and the precision of the
retrieval would drop significantly.

\todo{HSC: The following two sentences need to be merged and explained further.
Also, paragraphs need to have more than one sentence. Fix this.}
The basic assumption of our approach is that the two entities of a definition
relation co-occur in the same sentence.

The basic assumption of our approach is that if two entities take part in a
relation, those two co-occur in the same sentence.
Any other sentence is, therefor, irrelevant.

Another assumption can be made about the lexical class
of the description we want to rank. They are nouns or even noun phrases 
(e.g.,
\texttt{`the effective spring constant k'} or \texttt{`mass m of
something'}). We discarded all other words (according to their POS tag)
except noun phrases and Wikipedia hyperlinks.  These are candidates for a 
definition.  This is important, due to the fact that overall
ranking will be greatly influenced by the distance of candidate words. For all
intents and purposes, it is not necessary to keep noun phrases as single words.
They can therefore be aggregated and be ranked as if they were a single word.

@@

The classic tf-idf \cite{Salton86} statistic reflects the importance of a term
to a document. For our task, the inverse document frequency (idf) assigns
high penalties to frequent words like `length', as opposed to words seldom
seen such as `Hamiltonian'. These are both valid definitions for identifiers. 
As the influence of $\mathrm{tf}(t,s)$ on the sensitivity of the overall ranking
(\ref{eq:rating}) seems to be very high, we reduce the impact with the tuning
parameters $\gamma=0.1$ and $\alpha = \beta = 1$. Please note that the algorithm 
currently only takes sentences into account which were found in a single article. 
In the future, the retrieval process will examine sets of closely related articles.
This will leverage the problem that distributional properties will be volatile on 
term universes with very few members (e.g., term frequencies in a single sentence).

@@

\subsection{Description Retrieval}
We ran our program on a dataset of about 20,000 articles, all containing
\texttt{<math/>} tags, and retrieved about 550,000 candidate relations. The
most common definition relations were:

\begin{table}[H]
	\begin{center}
	\begin{tabular}{| l | p{6.8cm} | l |}
		\hline
		Identifier & Descriptions & Count\\
		\hline
		$n$ & number & 1709 \\
		$t$ & time & 1480 \\
		$M$ & mass & 1042 \\
		$r$ & radius & 752 \\
		$T$ & temperature & 666 \\
		$\theta$ & angle & 639 \\
		$G$ & group & 635 \\
		\hline
	\end{tabular}
	\end{center}
\caption{Most common definition relations}
\end{table}

\paragraph{Observations.} We observed some poorly ranked relations. For
example, in the fragment \texttt{`where $\phi$( $r_{i}$ ) is the electrostatic
potential'}, the distance $\Delta(\phi, \mathsf{electrostatic\:potential} ) = 6$. This is due
to counting brackets and function arguments as words. Also wrongly tagged
words like `Hamiltonian' as an adjective leads to false negatives. Latter
\todo{Latter ........ ???}

@@

There is no gold standard evaluation datasets to measure the
performance of identifier definition extractors. Thus created on on our own dataset.
This is a very time consuming job. At the moment, the dataset only contains two
large articles (revision ids are included) with around 100 identifier
definitions. This dataset is also available on the project repository.

@@

The distance measure $R_{\sigma_d}$ fails for the example of
Fig.~\ref{fig:screenshot} since $\Delta(\mathsf{energy},E) =
\Delta(\mathsf{mass},m) = 2$. Unfortunately, one cannot simply detect
punctuation marks and introduce some kind of directed associativity 
(e.g., inflicting a penalty on the ranking if the candidate relations spans over a
comma). This leads to whole classes of relation `types' (in terms of the
grammatical structure) never being retrieved. We plan to mitigate this problem
by taking more closely related scientific articles (based on their specific
fields) into consideration and count the frequencies of the candidate
relations. The intuition behind this is, that articles of the same scientific
field will likely use the same definition for the identifiers. Moreover,
we hope also to resolve the problem of `dangling' identifiers (those not
mentioned in the article itself), as there may be a description in related
articles.

@@

Our experiments showed that selecting candidates according to their POS tags
combined with numerical statistics about the text surface, can lead to quality
results. However, this approach is only applicable under certain conditions.
For identifiers which are seldom seen, our statistical approach tends to fail.
In that situation, other methods, especially supervised ones, are used to
bootstrap a language model. Unfortunately, they require a labeled test corpus
to measure the performance of a classifier that could be trained with our
generated data. Currently, we are planning to use the NTCIR-Math-10 Task, Math
Understanding Subtask gold standard dataset \cite{overview} for a comparable
evaluation.

@@

\todo{Fix two (?) dates in reference [3].}

>>>>

To build such a system, a labeled text corpus that annotates identifiers and
their definition is desirable. At the project start, such a corpus was not
available.
We calculate the probablities for correct identifier definition tuples based
on part-of-speech (POS) tagged based distance metrics.
\todo{rewrite sentences to make it process more clear}
 The motivation for our
choice is that readers usually extract identifier definitions from context
that is given by the surrounding text. In combination with the Stratosphere
system \cite{Alexandrov2010} that is capable of scalable data processing, we
rapidly generate a large amount of definition relation candidates with only
minimal implementation overhead for the parallelization.

As the target text corpus we chose the Wikipedia because of two facts. First,
most articles make use of \texttt{<math/>} tags (\emph{texvc} as an input
language) for formulae. The identification of these tags are easy to find,
and from the MathML output, it is easy to extract the identifiers.
Second, the text is already annotated with mark-up.
Particularly, hyperlinks to other articles within Wikipedia are of
interest as they typically wrap around any number of words and indicate that
these in combination are relevant in the given context or (respectively) sentence.

@@

At first, we implemented a simple identifier definition extractor that is
based on a set of static patterns. As this is a fairly robust approach and easy
to implement, it serves as a good reference point in terms of
performance. It simply iterates through the text, trying to find word groups,
that are matched by a pattern. The patterns being used to discover description
terms are depicted in table \ref{tpatterns}. Due to the fact that we already
tokenized an annotated the articles in a previous step in the MLP system, we
can make use of POS tags here as well.

Note, determiners not only contain articles, but also quantifiers and
distributives. The last pattern in table \ref{tpatterns} contains `*/DT'. This
is a shorthand for every word, that has the POS tag `DT' (determiner), will be
matched at this position. Otherwise this pattern whould be rather large, as it
needs to contain every possible determiner. \emph{identifer} as well as
\emph{description} are placholders, that mark the positions of the entities
from a possible defintion relation.

\begin{table}
\vspace{-5pt}
	\begin{center}
		\begin{tabular}{| p{9.3cm} |}
			\hline
			Pattern \\
			\hline
			\texttt{\emph{<description>} \emph{<identifier>}} \\
			\texttt{\emph{<identifier>} is \emph{<description>}} \\
			\texttt{\emph{<identifier>} is the \emph{<description>}} \\
			\texttt{let \emph{<identifier>} be the \emph{<description>}} \\
			\texttt{\emph{<description>} is|are denoted by \emph{<identifier>}} \\
			\texttt{\emph{<identifier>} denotes */DT \emph{<description>}} \\
			\hline
		\end{tabular}
	\end{center}
\caption{\label{tpatterns}Implemented static patterns}
\vspace{-5pt}
\end{table}

@@

Extracting relevant identifiers from the article rely on the assumption that
the author will use \texttt{<math/>} tags for all formulae. This said, a
formula that is written within the text cannot be recognized, and therefore
can never be extracted by our system.

The fact that we can estimate all relevant identifiers for an article (see
Section \ref{ir}), combined with some common assumptions about the relation,
can be exploited to largely reduce the set of candidates that need to be
ranked. Please note that this reduction is essential for retrieving the
correct relations for our approach. Otherwise almost any word would be ranked
and the precision of the retrieval would drop significantly.

The basic assumption of our approach is that the two entities of a definition
relation co-occur in the same sentence. In other words, if we want to retrieve
the description for an identifier, in the set of sentences that contain the
identifier the description should be found. Thus said, any other sentences can
be ignored. Furthermore, we assume that it is more likely to find the
description in first sentences than in the latter. This is based on the idea
that authors introduce the meaning of an identifer and than subsequent use it,
without necessarily repeating the meaning.

Another assumption can be made about the lexical class of the description we
want to rank. They are nouns or even noun phrases (e.g., \texttt{`the
effective spring constant k'} or \texttt{`mass m of something'}). We discard
all other words (according to their POS tag) except noun phrases and Wikipedia
hyperlinks. These are the candidates for a definition relation. Noun phrases
and hyperlinks may consist of multiple words. For all intents and purposes, it
is not necessary to threat noun phrases and hyperlinks as a set of words, and
therefore, will be threated subsequently as if they were one. This is
important, due to the fact that the overall ranking will be greatly influenced
by the distance of candidates to the position of the identifer.

@@

The classic tf-idf \cite{Salton86} statistic reflects the importance of a term
to a document. For our task, the inverse document frequency (idf) assigns
high penalties to frequent words like `length', as opposed to words seldom
seen such as `Hamiltonian'. These are both valid definitions for identifiers.
As the influence of $\mathrm{tf}(t,s)$ on the sensitivity of the overall ranking
(\ref{eq:rating}) seems to be very high, we reduce the impact with the tuning
parameters $\gamma=0.1$ and remain $\alpha = \beta = 1$. Please note that the algorithm
currently only takes sentences into account which were found in a single article.
In the future, the MLP system will examine sets of closely related articles.
This will leverage the problem that distributional properties will be volatile on
term universes with very few members (e.g., term frequencies in a single sentence).

@@

\subsection{Description Retrieval}
We ran our program on a dataset of about 20,000 articles, all containing
\texttt{<math/>} tags, and retrieved about 550,000 candidate relations. The
most common definition relations are listed in table \ref{tcommon}.

\begin{table}[H]
	\begin{center}
	\begin{tabular}{| l | p{6.8cm} | l |}
		\hline
		Identifier & Descriptions & Count\\
		\hline
		$n$ & number & 1709 \\
		$t$ & time & 1480 \\
		$M$ & mass & 1042 \\
		$r$ & radius & 752 \\
		$T$ & temperature & 666 \\
		$\theta$ & angle & 639 \\
		$G$ & group & 635 \\
		\hline
	\end{tabular}
	\end{center}
\caption{\label{tcommon}Most common definition relations}
\end{table}

\paragraph{Observations.} We observed some poorly ranked relations. For
example, in the fragment \texttt{`where $\phi$( $r_{i}$ ) is the electrostatic
potential'}, the distance $\Delta(\phi, \mathsf{electrostatic\:potential} ) = 6$. This is due
to counting brackets and function arguments as words. Also wrongly tagged
words like `Hamiltonian' as an adjective leads to false negatives.

@@

There is no gold standard evaluation datasets to measure the performance of
identifier definition extractors. Thus we created on on our own dataset. This
is a very time consuming job. At the moment, the dataset only contains two
large articles (revision ids are included) with around 100 identifier
definitions. This dataset is also available on the project repository.

@@

The distance measure $R_{\sigma_d}$ fails for the example of
Fig.~\ref{fig:screenshot} since $\Delta(\mathsf{energy},E) =
\Delta(\mathsf{mass},m) = 2$. Unfortunately, one cannot simply detect
punctuation marks and introduce some kind of directed associativity
(e.g., inflicting a penalty on the ranking if the candidate relations spans over a
comma). This leads to whole classes of relation `types' (in terms of the
grammatical structure) never being retrieved. We plan to mitigate this problem
by taking more closely related scientific articles (based on their specific
fields) into consideration and count the frequencies of the candidate
relations. The intuition behind this is, that articles of the same scientific
field will likely use the same definition for the identifiers. Moreover,
we also hope to resolve the problem of `dangling' identifiers (those not
mentioned in the article itself), as they might be described in related
articles.

@@

Our experiments showed that selecting candidates according to their POS tags
combined with numerical statistics about the text surface, can lead to quality
results. However, this approach is only applicable under certain conditions.
For identifiers which are seldom seen, our statistical approach tends to fail.
In that situation, other methods, especially supervised ones, are preferred.
Unfortunately, many of them require a labeled test corpus to measure the
performance of a classifier that could be trained with our generated data.
Currently, we are planning to use the NTCIR-Math-10 Task, Math Understanding
Subtask gold standard dataset \cite{overview} for a comparable evaluation.

@@
